---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---


## Loading and preprocessing the data

1. Load `data.table` and `dplyr` packages for later data manipulation, and
`ggplot2` plus `scales` for plotting.

```{r message=FALSE, warning=FALSE}
if (!require(data.table)) {install.packages("data.table"); library(data.table)}
if (!require(dplyr)) {install.packages("dplyr"); library(dplyr)}
if (!require(ggplot2)) {install.packages("ggplot2"); library(ggplot2)}
if (!require(scales)) {install.packages("scales"); library(scales)}
```

2. Download data file and extract the **.zip** file if not already present in 
**data/** directory
3. Read **.csv** file using `data.table::fread` wrapping the output using the 
`%>%` operator which pipes the output first into a `data.table` object, then a 
`dplyr::tbl_dt` object.


```{r} 
if (!file.exists("data/activity.csv")) {
    download.file(url = 
        "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", 
        method = "curl", destfile = "data/temp.zip", quiet = TRUE)
    unzip(zipfile = "data/temp.zip", exdir = "data")
    file.remove("temp.zip")
    }
activity <- fread("data//activity.csv") %>%
    data.table %>%
    tbl_dt 
```

Next we alter the loaded table using `dplyr`'s mutate function

1. Convert date to integer-based `IDate` class from `data.table` package. Though 
not useful for this smaller data table, it can sort more quickly with very large 
data sets
2. Likewise, `ITime` from the `interval` column. 
  - `Interval <integer division> 100` gives the hours
  -  `Interval <modulo> 100` gives the minutes 
3. `date, time` is created as a data.table key, thereby accelerating any future 
access of the data.

```{r}
activity <- activity %>%
    mutate(date=as.IDate(date),
           time=as.ITime(
               sprintf("%02d:%02d",
                       interval %/% 100,
                       interval  %% 100)))
setkey(x = activity, date, time)
```
Now we have 
```{r}
tables()
```

Which looks like
```{r}
head(activity)
```


## What is mean total number of steps taken per day?

To summarize the data into total steps taken per day, we could drop all time 
intervals that include an NA

```
activity %>% filter(!is.na(steps)) %>% group_by(date) %>% summarise(sum(steps))
```

However, that completely drops some dates, such as 2012-10-01 and 2012-11-01,
which have no valid measurements. At this point in the assignment, rather than 
imputing missing values, we are *trying* to show the limitations of ignoring 
the NAs. One way to do this is to read the NAs as zeros which simply do not add 
to the sum of steps.

```{r}
activity <- activity %>% 
    mutate(stepsZeroNAs = ifelse(test = is.na(steps), yes = 0, no = steps))
#stepsZeroNAs column = Steps with ZERO in place of NAs

stepCounts <- activity %>% 
    group_by(date) %>% 
    summarise(stepsZeroNAs=sum(stepsZeroNAs))
```

Now, lets build a histogram of the data using ggplot2
```{r}
# color
tempHue <- 0.58 
ColA.light <- hsv(h = tempHue, s = 0.4, v = 0.6, alpha = 0.4)
ColA.dark <- hsv(h = tempHue, s = 0.4, v = 0.4, alpha = 0.8)
# mean and median
meanstepsZeroNAs <- mean(stepCounts$stepsZeroNAs) 
medianstepsZeroNAs <- median(stepCounts$stepsZeroNAs)

dailySteps <- 
  # choose data
  ggplot(data = stepCounts) +
  # aesthetics, one series => histogram
  aes(x = stepsZeroNAs) +
  # geometric details of histogram, including coloring
  geom_histogram(fill=ColA.light, colour=ColA.dark, binwidth = 1000) +
  # annotate mean and median values
  geom_vline(xintercept=meanstepsZeroNAs, colour=ColA.dark, size=1.0, linetype="longdash") +
  geom_vline(xintercept=medianstepsZeroNAs, colour=ColA.dark, size=1.0, linetype="longdash") +
  # Add labels
  labs(
    title = "Distribution of Steps per Day", 
    x = "Daily Total Steps", y = "Days Count")

# execute the plot
dailySteps
```

**`r sprintf("%.0f", meanstepsZeroNAs)`** is the **mean** number of steps 
taken in a day, as calculated from data where NA is treated like no steps in 
that time period

**`r sprintf("%.0f", medianstepsZeroNAs)`** is the **median**, using the same data

## What is the average daily activity pattern?

Calculate typical daily pattern, with NAs ignored and with NAs treated as zero.
```{r}
dailyPattern <- activity %>%
    group_by(time) %>% 
    summarise(stepsZeroNAs=mean(stepsZeroNAs), 
              stepsTypical=mean(steps, na.rm = TRUE))
setkey(dailyPattern, time)
```
Then plot the data

```{r message=FALSE, warning=FALSE}
qplot(data = dailyPattern, y=stepsZeroNAs, x=time/3600, geom="line", 
      xlim = c(0,24), xlab="Time Interval (hour)", ylab="Steps", 
      main="Daily Pattern without Imputed Data")
```

From this calculation, we can note that typically, the most active 5-minute 
interval during the day starts at:
```{r}
dailyPattern[stepsZeroNAs==max(dailyPattern$stepsZeroNAs), time]
```

## Imputing missing values

The original data set contained **```r sum(is.na(activity$steps))```** NAs, 
out of a total **```r length(activity$steps)```** observations.

The typical number of steps at each time interval during the day was calculated 
in the previous section

```
dailyPattern <- activity %>%
    group_by(time) %>% 
    summarise(stepsZeroNAs=mean(stepsZeroNAs), 
              stepsTypical=mean(steps, na.rm = TRUE))
```

This table can be used to impute typical values for the NAs in the original 
data set. It covers all time intervals during the day covered by the earlier 
method of turing NAs into zeroes:
```{r}
length(unique(dailyPattern$stepsTypical)) == length(unique(dailyPattern$stepsZeroNAs))
```

And, it contains no NAs:
```{r}
sum(is.na(length(unique(dailyPattern$stepsTypical))))
```

This table can be used to create a new column in the original data set that 
imputes missing values from the typical daily pattern in cases there were 
NA values previously.

```{r}
activity <- activity %>% 
    mutate(stepsTypical = ifelse(
        test = is.na(steps), 
        yes = dailyPattern[time==time, stepsTypical],
        no = steps))
```

These new data can be plotted like before:

```{r}
# add imputed data
stepCounts <- activity %>% 
    group_by(date) %>% 
    summarise(stepsZeroNAs=sum(stepsZeroNAs),
              stepsImputed=sum(stepsTypical)) 

# mean and median
meanstepsImputed <- mean(stepCounts$stepsImputed)
medianstepsImputed <- median(stepCounts$stepsImputed) 

# new colors
tempHue <- 0.88
    ColB.light <- hsv(h = tempHue, s = 0.4, v = 0.6, alpha = 0.4)
    ColB.dark <- hsv(h = tempHue, s = 0.4, v = 0.4, alpha = 0.8)

# build the plot
dailySteps.Imputed <- 
  # choose data
  ggplot(data = stepCounts) +
  # aesthetics, one series => histogram
  aes(x = stepsImputed) +
  # geometric details of histogram, including coloring
  geom_histogram(fill=ColB.light, colour=ColB.dark, binwidth = 1000) +
  # annotate mean and median values
  geom_vline(xintercept=meanstepsImputed, colour=ColB.dark, size=1.0, linetype="longdash") +
  geom_vline(xintercept=medianstepsImputed, colour=ColB.dark, size=1.0, linetype="longdash") +
  # Add labels
  labs(
    title = "Distribution of Steps per Day (With Imputed Data)", 
    x = "Daily Total Steps", y = "Days Count")

# execute the plot
dailySteps.Imputed



```


## Are there differences in activity patterns between weekdays and weekends?

```{r fig.width=10, message=FALSE, warning=FALSE}
activity <- activity %>%
  mutate(weekday= ifelse(
    weekdays(date) %in% c("Saturday", "Sunday"), 
    "Weekend", 
    "Weekday"))

dailyPattern.split <- activity %>%
    group_by(time, weekday) %>%
    summarise(time, steps=mean(stepsTypical))

ggplot(data = dailyPattern.split, aes(x = as.POSIXct(time), y = steps, color = weekday)) + 
  scale_x_datetime(breaks = date_breaks("2 hour"), labels = date_format("%l %p")) +
  # See documentation for `scales` package to understand `scale_x_datetime`,
  # `date_breaks` and `date_format` transformation of times along x-axis
  facet_grid(weekday ~ .) + geom_line() + 
  labs(title = sprintf(
    "Distribution of Steps Taken \nby Time of Day \nWeekend vs. Weekdays"), 
    x = "Interval", y = "Steps per 5-minute Interval")
```

And, just for fun here is an additional plot of the data
```{r fig.width=10, fig.height=8, message=FALSE, warning=FALSE}
ggplot(data = activity, aes(x = as.POSIXct(time), y = stepsTypical, colour = weekday, fill = weekday)) + 
  scale_x_datetime(breaks = date_breaks("2 hour"), labels = date_format("%l %p")) +
  geom_smooth(level = 0.8) + geom_point(alpha=0.5, size=1, position = "jitter") +
  labs(title = sprintf(
    "Distribution of Steps Taken \nby Time of Day \nWeekend vs. Weekdays"), 
    x = "Time of Day", y = "Steps per 5-minute Interval")
```

And, zooming in on the smoothed data
```{r fig.width=10, fig.height=6, message=FALSE, warning=FALSE}
ggplot(data = activity, 
       aes(x = as.POSIXct(time), y = stepsTypical, colour = weekday, fill = weekday)) + 
  scale_x_datetime(breaks = date_breaks("2 hour"), labels = date_format("%l %p")) +
  geom_smooth(level = 0.6) +
  labs(title = sprintf("Distribution of Steps Taken \nby Time of Day \nWeekend vs. Weekdays"), 
       x = "Time of Day", y = "Steps per 5-minute Interval")
```
